String Reduction Schemas

### 2025-07-13 Paper/string-reduction-schemas.tex
DCFL <= 1-SR proof

We also define the \emph{stack function} $\S\colon\T^*\times\Gamma^*\partial\Gamma^*$ recursively as follows:
\begin{enumerate}
    \item $\S(\epsilon;W)=W$
    \item if $\bar t$ ends with $\pi_e(\bar t)=q$, then
    \begin{enumerate}
        \item $\S(\bar t(q\vdash p);W)=\S(\bar t;W)$
        \item $\S(\bar t(q\vdash p+W');W)=\S(\bar t;W'W)$
        \item $\S(\bar t(q\vdash p-A);AW)=\S(\bar t;W)$
    \end{enumerate}
\end{enumerate}
Otherwise $\S$ is left undefined.

@@

We define the \emph{transition operator} $\tr\colon\T\times\T\partial\T$ as follows:
\begin{align*}
    (p\vdash q+ A),\,(p\vdash r- A) &\mapsto (p\vdash r)\\
    (p\vdash q+ A),\,(q\vdash r) &\mapsto (p\vdash r+ A)\\
    (p\vdash q),\,(q\vdash r+ A) &\mapsto (p\vdash r+ A)\\
    %(p\vdash q- A),\,(q\vdash r) &\mapsto (p\vdash r- A)\\
    %(p\vdash q),\,(q\vdash r- A) &\mapsto (p\vdash r- A)\\
    (p\vdash q),\,(q\vdash r) &\mapsto (p\vdash r)
\end{align*}
This extends to $\tr\colon\P(\T)\times\P(\T)\to\P(\T)$ by $\tr(S_1,S_2)=\set{\tr(s_1,s_2)}[s_i\in S_i]$.

@@

We define a $\SR1$ schema $\widehat B$ which will accept $\L(\widehat P)$.
We define
$$ \widehat B = (\Sigma\cup\hep, \hat\Lambda, \square, \set\top, \hib) $$
where
$$ \hat\Lambda = \P(\T) \cup \P(Q\top) \cup \set{\square,\top} $$
where $Q\top=\set{p\top}[p\in Q]$.
And we define the initial reduction function as follows:
\begin{enumerate}
    \item $\hib(S_1,S_2)=\tr(S_1,S_2)$ for $S_i\in\P(\T)$ if this is non-empty; otherwise leave it undefined.
    \item $\hib(S,\square)=\set{p\top}[(p\vdash f+ A)\in S\hbox{ or }(p\vdash f)\in S,\hbox{ where $f\in F$}]$ for $S\in\P(\T)$, if this is non-empty; otherwise leave it undefined.
    \item $\hib(S_1,S_2)=\set{p\top}[(p\vdash q+ A)\in S_1\hbox{ or }(p\vdash q)\in S_1\hbox{ and }q\top\in S_2]$ for $S_1\in\P(\T)$ and $S_2\in\P(Q\top)$, if this is non-empty; otherwise leave it undefined.
    \item $\hib(\square,S)=\top$ if $S\in\P(Q\top)$ and $q_0\top\in S$.
\end{enumerate}

@@

    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\square\hb^*(\bar S)\square$.

@@

    $\square,S$ is only reduced if $S\in\P(Q\top)$.
    And in general we have that if $\bar S\hto^*\bar T$ then $\bar S\square\hto^*\bar T\square$.
    And since $\square\bar S\hto^*\square\hb^*(\bar S)$ (since reducing $\bar S$ does not produce elements of $\P(Q\top)$), we get the desired result.
    \qed

@@

    For $\bar S\in\P(\T)^*$, let $\bar T=\hb^*(\bar S)$.
    \begin{enumerate}
        \item if $\bar T=T$, then for $t\in T$ if $\S(t)$ is defined then $t$ is not of the form $p\vdash q- A$.
        \item otherwise, for any $t_i\in T_i$ for which $\S(\bar t)$ is defined, $t_i$ is of the form $p\vdash q+ A$.
    \end{enumerate}

@@

    $\S(p\vdash q- A)$ is undefined, so the first point is trivial.

    Now if $\bar T$ has a length greater than $1$, let $i$ be the first index where $p\vdash q=t_i$ or $p\vdash q- A=t_i$.
    If $p\vdash q=t_i$, then if $i=1$, we can look to $t_{i+1}$, which must be either $q\vdash r$ or $q\vdash r+ A$, and in any case $\tr$ is defined and so $\bar T$ can be reduced, in contradiction.
    If $i>1$, then similarly we look to the left of $i$, and this must be of the form $r\vdash p+ A$ (since $i$ is the first index in which something other than $p\vdash q+ A$ occurs), and again $\tr$ is defined.

    Similarly if $p\vdash q- A=t_i$ then $i>1$, and in order for the stack to be defined $t_{i-1}=r\vdash p+ A$, so $\tr$ can be defined.
    \qed

@@

    For $\bar S\in\P(\T)^*$, $\S(\bar S)=\S(\hb(\bar S))$

This is immediate since $\hb$ only applies $\tr$ on $\bar S$, which doesn't affect $\S$.

    For $\bar S\in\P(\T)^*$, $\square\bar S\square\hto^*\top$ if and only if $(q_0,W,f)\in \S(\bar S)$.

@@

    Let $\bar T=\hb^*(\bar S)$, then we know that $\square\bar S\square\hto^*\square\bar T\square$ and $\S(\bar S)=\S(\bar T)$.

    For $\bar T=T$, then $\square T\square\hto^*\top$ if and only if $q_0\vdash f$ or $q_0\vdash f+ A$ is in $T$.
    And in either case $(q_0,W,f)\in\S(\bar T)$.
    Now, if $\bar T=\bar T'T$, then $\square\bar T'T\square\hto^*\top$ means that $p\vdash f+ A\in T$ so that it can reduce with $\square$.
    Then $\bar T'$ reduces with $\hib(T,\square)$ if and only if $q\vdash p+ A$ is in the end of $\bar T'$ and $p\vdash f+ B$ is in $T$.
    And so on inductively we see that this means that $\bar T$ starts with $q_0\vdash q_1+ A_0$ and then has $q_1\vdash q_2+ A_1$ and so on until $q_n\vdash f+ A_n$.
    Thus $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$.

    And conversely if $(q_0,A_0\cdots A_n,f)\in\S(\bar T)=\S(\bar S)$, then this means that $\bar T$ consists of $T_0\cdots T_n$ such that $q_i\vdash q_{i+1}+ A_i\in T_i$.
    Thus $\square\bar T\square\hto^*\top$ as required.
    \qed

@@

@@

    The following are equivalent:
    \begin{enumerate}
        \item $w\in\L(P)$
        \item $\epsilon(w)\in\L(\hat B)$
        \item there exists a $\hat w\in E(w)$ such that $\hat w\in\L(\hat B)$
    \end{enumerate}

@@

    We know that for $\hat w\in E(w)$, $\hat w\in\L(\hat B)$ if and only if $\square\uline{\hat w}\square\hto^*\top$ iff $(q_0,W,f)\in\S(\uline{\hat w})$ iff $\hat w\in\L(\hat P)$.
    Since this is for an arbitrary $\hat w\in E(w)$, this proves the three equivalences.
    \qed

Now we claim that $w\in\L(B)$ iff $\epsilon(w)\in\L(\hat B)$.

>>>>

We also define the \emph{stack function} $\S\colon\T^*\times\Gamma^*\partial\Gamma^*$ recursively as follows:
\begin{enumerate}
    \item $\S(\epsilon;W)=W$
    \item if $\bar t$ ends with $\pi_e(\bar t)=q$, then
    \begin{enumerate}
        \item $\S(\bar t(q\vdash p);W)=\S(\bar t;W)$
        \item $\S(\bar t(q\vdash p+A);W)=A\S(\bar t;W)$
        \item if $\S(\bar t;W)=AW'$ then $\S(\bar t(q\vdash p-A);W)=W'$
    \end{enumerate}
\end{enumerate}
Otherwise $\S$ is left undefined.

@@

We define the \emph{transition reduction operator} $\tr\colon\T\times\T\partial\T$ as follows:
\begin{align*}
    (p\vdash q+ A),\,(p\vdash r- A) &\mapsto (p\vdash r)\\
    (p\vdash q+ A),\,(q\vdash r) &\mapsto (p\vdash r+ A)\\
    (p\vdash q),\,(q\vdash r+ A) &\mapsto (p\vdash r+ A)\\
    %(p\vdash q- A),\,(q\vdash r) &\mapsto (p\vdash r- A)\\
    %(p\vdash q),\,(q\vdash r- A) &\mapsto (p\vdash r- A)\\
    (p\vdash q),\,(q\vdash r) &\mapsto (p\vdash r)
\end{align*}
This extends to $\tr\colon\P(\T)\times\P(\T)\to\P(\T)$ by $\tr(S_1,S_2)=\set{\tr(s_1,s_2)}[s_i\in S_i]$.

@@

We define $\S_\epsilon$ similar to $\S$, accept we allow $\epsilon$-transitions.
Specifically, $\S_\epsilon\colon Q\times\T^*\times\Gamma^*\partial\P_{<\omega}(\Gamma^*\times Q)$ where first we define
$$ \S_\e(p_0,\e;W)=\set{(W',q)}[(p_0,\e,W)\vdash^*(q,\e,W')] $$
and we recurse:
\begin{enumerate}
    \item for $(p\vdash q)$, if $(W',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,W')\vdash^*(s,\epsilon,W'')$ then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q);W)$.
    \item for $(p\vdash q+A)$, if $(W',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,AW')\vdash^*(s,\epsilon,W'')$, then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q+A);W)$.
    \item for $(p\vdash q-A)$, if $(AW',p)\in\S_\e(p_0,\bar t;W)$ and $(q,\epsilon,W')\vdash^*(s,\epsilon,W'')$, then $(W'',s)\in\S_\e(p_0,\bar t(p\vdash q+A);W)$.
\end{enumerate}
We extend $\S_\e$ to $\P_{<\omega}(\T)^*\times\Gamma^*\to\P_{<\omega}(Q\times\Gamma^*\times Q)$ where
$$ \S_\e(\bar S;W) = \set{(p_0,W',p_e)}[p_0\in Q,\,(W',p_e)\in\S_\e(p_0,\bar S;W)] $$

Furthermore, for $\bar S=S_1\cdots S_k\in\P(\T)^*$, let us define
$$ \uline E(\bar S) = \set{\uline\hep^{n_0}S_1\uline\hep^{n_1}\cdots S_k\uline\hep^{n_k}}[n_i\geq0] $$
so that $\uline E(\uline w)=\uline{E(w)}$.

An immediate result by definition is given two strings of the same length $\bar S$ and $\bar T$, if we define their union $\bar S\cup\bar T$ pointwise (meaning $(\bar S\cup\bar T)_i=S_i\cup T_i$), then we have
$$ \S_\e(\bar S\cup\bar T;W) = \S_\e(\bar S;W)\cup\S_\e(\bar T;W) $$

@@

    $$ \S_\e(\bar S;W) = \bigcup_{\hbar S\in\uline E(\bar S)}\S(\hbar S;W) $$
    in particular
    $$ \S_\e(\uline w;\e) = \bigcup_{\hat w\in E(w)}\S(\hat{\uline w};\e) $$

@@

    This is done via induction on $\bar S$.
    For $\bar S=\e$, if $(p_0,W',p_e)\in\S_\e(\e;W)$ then $(p_0,\e,W)\vdash^n(p_e,\e,W')$ then $(p_0,W',p_e)\in\S(\hep^n;W)$.
    And so on, the idea being that when $(q,\e,W')\vdash^n(s,\e,W'')$, we replace this with $\hep^n$.
    \qed

@@

    For $w\in\Sigma^*$,
    $$ (p,W',q)\in\S_\e(\uline w;W) \iff (p,w,W)\vdash^*(q,\e,W') $$

@@

    This is immediate due to the previous lemma and lemma \ref{lem:stackandpda}.

    For $\bar S\in\P(\T)^*$, $\S_\e(\bar S)=\S_\e(\beta\bar S)$.

    We must split into two cases: the first when only a single symbol of $\bar S$ is reduced, and the second when two symbols are reduced.

    In the first case, suppose $\bar S=S_1\cdots S_i\cdots S_n$, and $\beta\bar S=S_1\cdots\ib S_i\cdots S_n$.
    Now we know that $\ib(S_i)=S_i\cup\tr(S_i,\uline\hep)\cup\tr(\uline\hep,S_i)$, thus
    $$ \S_\e(\beta\bar S) = \S_\e(\bar S)\cup\S_\e(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)\cup\S_\e(S_1\cdots\tr(\uline\hep,S_i)\cdots S_n) $$
    We see then that $\S_\e(\bar S)\subseteq\S_\e(\beta\bar S)$.
    Conversely, let us look at $E_1=\uline E(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)$.
    Since $\S$ is $\tr$-invariant, taking $\S$ of $E_1$ is equal to $\S$ over $\uline E(S_1\cdots S_i\uline\hep\cdots S_n)\subseteq E(\bar S)$.
    Thus we see that $\S_\e(S_1\cdots\tr(S_i,\uline\hep)\cdots S_n)\subseteq\S_e(\bar S)$, and similar for $\tr(\uline\hep,S_i)$, as required.

    For the second case, this is just due to $\S$ being $\tr$-invariant.
    \qed

@@

    For $\bar S\in\P(\T)^*$, $\bar S\square\to^*\top\in\A$ if and only if $(q_0,W,f)\in \S_\e(\bar S;\e)$.

@@

    Let $\bar T=\hb^*(\bar S)$, then we know that $\bar S\square\to^*\bar T\square$ and $\S_\e(\bar S)=\S_\e(\bar T)$.

    Now, we know that $\bar T\square\to^*\top\in\A$ if and only if $p_i\vdash q_i+A_i\in T_i$ such that
    \begin{enumerate}
        \item $A_i\in\Gamma_\e$
        \item $q_i\vdash^* p_{i+1}$
        \item $q_0\vdash^* p_1$ and $q_n\vdash^* f\in F$
    \end{enumerate}
    And it is not hard to see that this is equivalent.
    \qed

@@

In particular we have that $w\in\L(B)$ if and only if $(q_0,W,f)\in\S_\e(\uline w;\e)$ if and only if $(q_0,w,\e)\vdash^*(f,\e,W')$ if and only if $w\in\L(P)$ as required.

@@

    $B$ and $P$ are equivalent.

@@

    $\DCFL\subseteq\SR1$
